For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-04 paper/paper.tex
use signature algebra instead of model

%% We define IIT signatures using the Church encoding introduced by
%% Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A
%% difference is that we avoid the usage of impredicativity because it is
%% not a problem for our use case that signatures are not in the lowest
%% universe.
\begin{definition}[IIT signature]\label{defn:sign}
  An IIT signature is a context in an arbitrary signature algebra, which is also
  compatible with morphisms:
  \begin{alignat*}{6}
    & \Sign :=\,\, && \big({sig}:(M:\SignAlg)\ra\Con^M\big)\times \\
    & && \big((M\,N:\SignAlg)(f:\SignMor\,M\,N)\ra f_\Con\,({sig}\,M) = {sig}\,N \big).
  \end{alignat*}
\end{definition}
The compatibility condition says that if we obtain an $M$-context
using ${sig}$ in model $M$ and then we transport it to $N$ using $f$,
we get the same $N$-context as directly applying ${sig}$ to $N$.

@@

\begin{theorem}\label{thm:ToSToIITs}
  If a model of type theory supports the theory of IIT signatures,
  then it supports all IITs.
\end{theorem}
\begin{proof}
  For a signature $({sig},{sigc})$, we define constructors as
  \[
    \con_{sig}:= (\ll{sig}\,\I\rr_{\IC_{{sig}\,\I}})^\C\,\id^\I : ({sig}\,\ADS)^\A
  \]
  This typechecks as
  $\ll{sig}\,\I\rr_\A = \ll\blank\rr_\A\,({sig}\,\I)
  \overset{{sigc}}{=} {{sig}\,\A} = ({sig}\,\ADS)^\A$. We define the
  eliminator by
  and an
  \[
    \elim_{sig}\,\gamma^D := (\ll{sig}\,\I\rr_{\IE_{\gamma^D}})^\E\,\id^\I  : ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
  This typechecks firstly because the type of $\gamma^D$ matches the
  type of the parameter of $\IE$:
  \[
    ({sig}\,\ADS)^\D\,\con_{sig} \overset{{sigc}}{=}
    (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\D\,\con_{sig} =
    (\ll{sig}\,\I\rr_{\ADS})^\D\,\con_{sig},
  \]
  and the result also has the correct type:
  \[
    \ll{sig}\,\I\rr^\S\,\con_{sig}\,\gamma^D = (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\S\,\con_{sig}\,\gamma^D \overset{{sigc}}{=} ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
\end{proof}

>>>>

%% We define IIT signatures using the Church encoding introduced by
%% Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A
%% difference is that we avoid the usage of impredicativity because it is
%% not a problem for our use case that signatures are not in the lowest
%% universe.
\begin{definition}[IIT signature]\label{defn:sign}
  An IIT signature is a context in an arbitrary signature algebra, which is also
  compatible with morphisms:
  \begin{alignat*}{6}
    & \Sign :=\,\, && \big({sig}:(M:\SignAlg)\ra\Con^M\big)\times \\
    & && \big((M\,N:\SignAlg)(f:\SignMor\,M\,N)\ra f_\Con\,({sig}\,M) = {sig}\,N \big).
  \end{alignat*}
\end{definition}
The compatibility condition says that if we obtain an $M$-context
using ${sig}$ at signature algebra $M$ and then we transport it to $N$ using $f$,
we get the same $N$-context as directly applying ${sig}$ to $N$.

@@

\begin{theorem}\label{thm:ToSToIITs}
  If a model of ETT supports the theory of IIT signatures,
  then it supports all IITs.
\end{theorem}
\begin{proof}
  For a signature $({sig},{sigc})$, we define constructors as
  \[
    \con_{sig}:= (\ll{sig}\,\I\rr_{\IC_{{sig}\,\I}})^\C\,\id^\I : ({sig}\,\ADS)^\A
  \]
  This typechecks as
  $\ll{sig}\,\I\rr_\A = \ll\blank\rr_\A\,({sig}\,\I)
  \overset{{sigc}}{=} {{sig}\,\A} = ({sig}\,\ADS)^\A$. We define the
  eliminator by
  and an
  \[
    \elim_{sig}\,\gamma^D := (\ll{sig}\,\I\rr_{\IE_{\gamma^D}})^\E\,\id^\I  : ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
  This typechecks firstly because the type of $\gamma^D$ matches the
  type of the parameter of $\IE$:
  \[
    ({sig}\,\ADS)^\D\,\con_{sig} \overset{{sigc}}{=}
    (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\D\,\con_{sig} =
    (\ll{sig}\,\I\rr_{\ADS})^\D\,\con_{sig},
  \]
  and the result also has the correct type:
  \[
    \ll{sig}\,\I\rr^\S\,\con_{sig}\,\gamma^D = (\ll\blank\rr_{\ADS}\,({sig}\,\I))^\S\,\con_{sig}\,\gamma^D \overset{{sigc}}{=} ({sig}\,\ADS)^\S\,\con_{sig}\,\gamma^D.
  \]
\end{proof}

### 2020-05-04 paper/universaliit.tex
finish
small fixes
fixes
use signature algebra instead of model

We construct the theory of IIT signatures in the following steps:
\begin{enumerate}
\item We view the theory of signatures as a type theory, and we define
  its untyped syntax as mutual inductive types together with typing
  judgments given by inductive relations on the untyped syntax. Then
  the syntax $\I : \SignAlg$ is constructed using those untyped terms
  for which the typing relation holds.
\item We construct $\ll\blank\rr:\SignMor\,\I\,M$ for arbitrary
  $M : \SignAlg$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\GG$ and a
    semantic context $\model{\GG}$ of the model $M$, we have
    $\GG \sim \model{\GG}$ if and only if $\ll\GG\rr
    = \model{\GG}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\GG}$ of
    % the model $M$ if and only if $\GG$ relates to $\rec\,{\GG}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional and thus obtaining a function.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\SignMor\,\I\,M$ satisfies the relation. For example, for any syntactic context
    $\GG$ we have $\GG\sim f\,\GG$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.

@@

Crucially, we do not have conversion relations for typed syntax, nor
do we need to use quotients when giving the syntactic model. This is
possible because there are no $\beta$-rules in the theory of
signatures. Hence, we consider only normal terms in the untyped
syntax, and define weakening and substitution by recursion.  Avoiding
quotients is important for two reasons. First, it greatly simplifies
formalisation. Second, we aim to reduce the theory of signatures only
to inductive types, thus making \Cref{thm:WToToS} stronger.

@@

\subsubsection{The functional relation}

@@

\subsubsection{Right Uniqueness}

@@

\begin{remark}
  We mentioned that in order to avoid a recursive-recursive definition, we
  removed some hypotheses in the list of arguments of the relation. Such
  hypotheses are sometimes missed, for example in the case of the empty
  substitution or in the case of variables, requiring us to state additional
  equalities. Because of this, we need UIP to show that
  $\sum_{\model{\GG}}\GG\sim\model{\GG}$ and
  $\sum_{\model{A}}A\sim\model{A}$ are propositions.  One may think that the use
  of UIP could be avoided by using the alternative verbose definition that we
  suggested before, expecting that
  $\sum_{\model{\GG}}\sum_{\model{A}}A\sim\model{A}$, rather than
  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.  However, this is not
  obvious. For example, we were not able to define
  $\Elw\,\wf{\GG}\,\wf{a}\sim \model{A}$ in this fashion. In related work,
  Hugunin investigated constructing IITs without
  UIP \cite{jasper} in cubical type theory, and demonstrated
  that well-formedness predicates used in syntactic models can subtly break
  in that setting. Also, while Hugunin does not use UIP, he only shows the simple version
  version of dependent elimination for the constructed IITs. Hence, the
  question whether IITs are reducible to inductive types in a UIP-free setting
  remains open.
\end{remark}

>>>>

We construct the theory of IIT signatures in the following steps:
\begin{enumerate}
\item We view the theory of signatures as a type theory, and we define
  its untyped syntax as mutual inductive types together with typing
  judgments given by inductive relations on the untyped syntax. Then
  the syntax $\I : \SignAlg$ is constructed using those untyped terms
  for which the typing relation holds.
\item We construct $\ll\blank\rr:\SignMor\,\I\,M$ for arbitrary
  $M : \SignAlg$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given signature algebra. The idea is that given a syntactic context $\GG$ and a
    semantic context $\model{\GG}$ of the signature algebra $M$, we have
    $\GG \sim \model{\GG}$ if and only if $\ll\GG\rr
    = \model{\GG}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\GG}$ of
    % the model $M$ if and only if $\GG$ relates to $\rec\,{\GG}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional and thus obtaining a morphism.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\SignMor\,\I\,M$ satisfies the relation. For example, for any syntactic context
    $\GG$ we have $\GG\sim f\,\GG$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.

@@

Crucially, we do not have conversion relations for typed syntax, nor
do we need to use quotients when constructing the syntax. This is
possible because there are no $\beta$-rules in the theory of
signatures. Hence, we consider only normal terms in the untyped
syntax, and define weakening and substitution by recursion.  Avoiding
quotients is important for two reasons. First, it greatly simplifies
formalisation. Second, we aim to reduce the theory of signatures only
to inductive types, thus making \Cref{thm:WToToS} stronger.

@@

\subsubsection{The Functional Relation}

@@

\subsubsection{Right Uniqueness}\label{sec:right_uniqueness}

@@

\begin{remark}
  We mentioned that in order to avoid a recursive-recursive definition, we
  removed some hypotheses in the list of arguments of the relation. Such
  hypotheses are sometimes missed, for example in the case of the empty
  substitution or in the case of variables, requiring us to state additional
  equalities. Because of this, we need UIP to show that
  $\sum_{\model{\GG}}\GG\sim\model{\GG}$ and
  $\sum_{\model{A}}A\sim\model{A}$ are propositions.  One may think that the use
  of UIP could be avoided by using the alternative verbose definition that we
  suggested before, expecting that
  $\sum_{\model{\GG}}\sum_{\model{A}}A\sim\model{A}$, rather than
  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.  However, this is not
  obvious. For example, we were not able to define
  $\Elw\,\wf{\GG}\,\wf{a}\sim \model{A}$ in this fashion. In related work,
  Hugunin investigated constructing IITs without
  UIP \cite{jasper} in cubical type theory, and demonstrated
  that well-formedness predicates used in syntactic algebras can subtly break
  in that setting. Also, while Hugunin does not use UIP, he only shows the simple version
  version of dependent elimination for the constructed IITs. Hence, the
  question whether IITs are reducible to inductive types in a UIP-free setting
  remains open.
\end{remark}
